Accuracy in Parameter Estimation and Simulation Approaches for Sample Size Planning with Multiple Stimuli

### 2023-12-28 manuscript_draft.Rmd
my updates from feedback

```{r trackdown, eval = F}
library(trackdown)
# initial set up 
# upload_file(file = "manuscript_draft.Rmd",
#       gfile = "power2stim_V2_trackdown",
#       gpath = "Studies/PSA007 SPAM-L/Manuscripts/Power",
#       shared_drive = "Psychological Science Accelerator")
# from google 
download_file(file = "manuscript_draft.Rmd", 
      gfile = "power2stim_V2_trackdown",
      gpath = "Studies/PSA007 SPAM-L/Manuscripts/Power",
      shared_drive = "Psychological Science Accelerator")
# to google
update_file(file = "manuscript_draft.Rmd", 
      gfile = "power2stim_V2_trackdown",
      gpath = "Studies/PSA007 SPAM-L/Manuscripts/Power",
      shared_drive = "Psychological Science Accelerator",
      rich_text = TRUE,
      hide_code = TRUE)
```

@@

These steps are summarized in Table \@ref(tab:table-summary). We will first demonstrate the ideas behind the steps using open data [@brysbaert2014; @balota2007]. This example will reveal a few areas of needed exploration for the steps. Next, we provide simulations for the proposed procedure to find solutions to overcome these potential limitations and formally define Step 2. Finally, we include additional resources for researchers to use to implement the proposed sample size estimation procedure.

@@

# grab only the overlap for this fake study
elpDF <- elpDF %>% 
  filter(Stimulus %in% conDF$Word)

@@

```{r con-sample, include = F, echo = F}
# sequence of sample sizes to try
samplesize_values <- seq(20, 300, 5)
# create a blank table for us to save the values in 
sim_table <- matrix(NA, 
          nrow = length(samplesize_values), 
          ncol = length(unique(con_use$Word)))
# make it a data frame
sim_table <- as.data.frame(sim_table)
# add a place for sample size values 
sim_table$sample_size <- NA
# loop over sample sizes
for (i in 1:length(samplesize_values)){

@@

 colnames(sim_table)[1:length(unique(con_use$Word))] <- temp$Word
 sim_table[i, 1:length(unique(con_use$Word))] <- temp$se
 sim_table[i, "sample_size"] <- samplesize_values[i]
}

# figure out potential samples
final_sample_c <- 
 sim_table %>% 
 pivot_longer(cols = -c(sample_size)) %>% 
 rename(item = name, se = value) %>% 
 group_by(sample_size) %>% 
 summarize(Percent_Below10 = sum(se <= cutoff_c["10%"])/length(unique(con_use$Word)),
          Percent_Below20 = sum(se <= cutoff_c["20%"])/length(unique(con_use$Word)),
          Percent_Below30 = sum(se <= cutoff_c["30%"])/length(unique(con_use$Word)),
          Percent_Below40 = sum(se <= cutoff_c["40%"])/length(unique(con_use$Word)),
          Percent_Below50 = sum(se <= cutoff_c["50%"])/length(unique(con_use$Word)),
          Percent_Below60 = sum(se <= cutoff_c["60%"])/length(unique(con_use$Word)),
          Percent_Below70 = sum(se <= cutoff_c["70%"])/length(unique(con_use$Word)),
          Percent_Below80 = sum(se <= cutoff_c["80%"])/length(unique(con_use$Word)),
          Percent_Below90 = sum(se <= cutoff_c["90%"])/length(unique(con_use$Word))) %>% 
  pivot_longer(-sample_size) %>% 
  filter(value >= .80) %>% 
  group_by(name) %>% 
  arrange(sample_size, value)

@@

```{r elp-sample, include = F, echo = F}
# sequence of sample sizes to try
samplesize_values <- seq(20, 300, 5)
# create a blank table for us to save the values in 
sim_table <- matrix(NA, 
          nrow = length(samplesize_values), 
          ncol = length(unique(elp_use$Stimulus)))
# make it a data frame
sim_table <- as.data.frame(sim_table)
# add a place for sample size values 
sim_table$sample_size <- NA
# loop over sample sizes
for (i in 1:length(samplesize_values)){

@@

 colnames(sim_table)[1:length(unique(elp_use$Stimulus))] <- temp$Stimulus
 sim_table[i, 1:length(unique(elp_use$Stimulus))] <- temp$se
 sim_table[i, "sample_size"] <- samplesize_values[i]
 }
# figure out potential samples
final_sample_e <- 
 sim_table %>% 
 pivot_longer(cols = -c(sample_size)) %>% 
 rename(item = name, se = value) %>% 
 group_by(sample_size) %>% 
 summarize(Percent_Below10 = sum(se <= cutoff_e["10%"])/length(unique(elp_use$Stimulus)),
          Percent_Below20 = sum(se <= cutoff_e["20%"])/length(unique(elp_use$Stimulus)),
          Percent_Below30 = sum(se <= cutoff_e["30%"])/length(unique(elp_use$Stimulus)),
          Percent_Below40 = sum(se <= cutoff_e["40%"])/length(unique(elp_use$Stimulus)),
          Percent_Below50 = sum(se <= cutoff_e["50%"])/length(unique(elp_use$Stimulus)),
          Percent_Below60 = sum(se <= cutoff_e["60%"])/length(unique(elp_use$Stimulus)),
          Percent_Below70 = sum(se <= cutoff_e["70%"])/length(unique(elp_use$Stimulus)),
          Percent_Below80 = sum(se <= cutoff_e["80%"])/length(unique(elp_use$Stimulus)),
          Percent_Below90 = sum(se <= cutoff_e["90%"])/length(unique(elp_use$Stimulus))) %>% 
  pivot_longer(-sample_size) %>% 
  filter(value >= .80) %>% 
  group_by(name) %>% 
  arrange(sample_size, value)

@@

results_table_example <- as.data.frame(results_table)
colnames(results_table_example) <- c("Deciles", "C Decile SE", "C 80", "C 85", "C 90", "C 95",
             "L Decile SE", "L 80", "L 85", "L 90", "L 95")
results_table_example$`C Decile SE` <- apa_num(as.numeric(results_table_example$`C Decile SE`), digits = 2)
results_table_example$`L Decile SE` <- apa_num(as.numeric(results_table_example$`L Decile SE`), digits = 2)

apa_table(results_table_example, 
          caption = "Sample Size Estimates by Decile for Concreteness Example",
          note = "Esimates are based on meeting at least the minimum percent of items (e.g., 80%) but may be estimated over that amount (e.g., 82.5%).")
```

@@

```{r scale-size-figure, echo = F, fig.cap="Simulated pilot sample size and final projected sample size to achieve 80%, 85%, 90%, and 95% of items below threshold. These values are averaged over all other variables. Black dots represent original sample size for reference. "}
likert_var <- sqrt(((7-1)^2)/4)
percent_var <- sqrt(((100-0)^2)/4)
millisecond_var <- sqrt(((3000-0)^2)/4)

@@

```{r scale-skew-figure, echo = F, fig.cap="Simulated pilot sample size compared to projected sample size for ceiling versus normal distributions on each scale. " }
ggplot(summary_long %>% 
         group_by(original_n, power, data_type) %>% 
         summarize(sample_size_avg = mean(sample_size)),
       aes(original_n, sample_size_avg, color = data_type)) + 
  geom_point() + 
  geom_point(aes(original_n, original_n), color = "black") + 
  geom_line() + 
  geom_line(aes(original_n, original_n), color = "black") + 
  theme_linedraw(base_size = 15) + 
  xlab("\nPilot Sample Size") + 
  ylab("Projected Sample Size\n") + 
  scale_color_manual(name = "Scale Skew",
                     values = c(plot_colors)) +
  facet_wrap(~power,
             labeller = labeller(power = power.labs))+
  theme_bw()
```

@@

```{r scale-hetero-figure, echo = F, fig.cap="Simulated pilot sample size compared to projected sample size for differing amounts of heterogeneity on each scale. "}
ggplot(summary_long %>% 
         group_by(original_n, power, variability) %>% 
         summarize(sample_size_avg = mean(sample_size)),
       aes(original_n, sample_size_avg, color = variability)) + 
  geom_point() + 
  geom_point(aes(original_n, original_n), color = "black") + 
  geom_line() + 
  theme_linedraw() + 
  xlab("Pilot Sample Size") + 
  ylab("Projected Sample Size") + 
  scale_color_discrete(name = "Item Heterogeneity") +
  facet_wrap(~power)
```

@@

```{r corrected-figure, echo = F, fig.cap="Corrected projected sample sizes for variability and power levels. "}
decay <- 1-sqrt((summary_long$original_n-20)/summary_long$original_n)^log2(summary_long$original_n)
summary_long$new_sample <- summary_long$sample_size*decay
ggplot(summary_long %>% 
         group_by(original_n, power, variability) %>% 
         summarize(sample_size_avg = mean(sample_size),
                   new_sample_size_avg = mean(new_sample)),
       aes(original_n, new_sample_size_avg, color = variability)) + 
  geom_line(aes(original_n, original_n), color = "black") +
  geom_point(aes(original_n, original_n), color = "black") + 
  geom_point() + 
  geom_line() + 
  xlab("\nPilot Sample Size") + 
  ylab("Projected Sample Size\n") + 
  scale_color_manual(name = "Item Heterogeneity",
                     values = c(plot_colors)) +
  facet_wrap(~power,
             labeller = labeller(power = power.labs))+
  theme_bw()
```

@@

The updated proposal steps are in Table \@ref(tab:table-summary). The main change occurs in Step 2 with a designated cutoff decile, and Step 5 with a correction score. Using the data from the 40% decile in Table \@ref(tab:table-example), we can determine that the stopping rule for concreteness ratings would be `r apa_num(cutoff_c[4])`, and the stopping rule for lexical decision times would be `r apa_num(cutoff_e[4])`. For Step 5, we apply our correction formula separately for each one, as they have different variability scores, and these scores are shown in Table \@ref(tab:table-scores-updated). Each row was multiplied by row one's formula, and then these scores are summed for the final sample size. Sample sizes cannot be proportional, so we recommend rounding up to the nearest whole number.

@@

```{r boot-sample, include = T, echo = T}
samples <- bootstrap_samples(start = 20, # starting sample size
  stop = 100, # stopping sample size
  increase = 5, # increase bootstrapped samples by this amount
  population = pops, # population or pilot data
  replace = TRUE, # bootstrap with replacement? 
  grouping_items = "item") # item column label

>>>>

```{r trackdown, eval = F}
# devtools::install_github("ClaudioZandonella/trackdown")
library(trackdown)
# initial set up 
# upload_file(file = "manuscript_draft.Rmd",
#       gfile = "power2stim_V2_trackdown",
#       gpath = "Studies/PSA007 SPAM-L/Manuscripts/Power",
#       shared_drive = "Psychological Science Accelerator")
# from google 
download_file(file = "manuscript_draft.Rmd", 
      gfile = "power2stim_V2_trackdown",
      gpath = "Studies/PSA007 SPAM-L/Manuscripts/Power",
      shared_drive = "Psychological Science Accelerator")
# to google
update_file(file = "manuscript_draft.Rmd", 
      gfile = "power2stim_V2_trackdown",
      gpath = "Studies/PSA007 SPAM-L/Manuscripts/Power",
      shared_drive = "Psychological Science Accelerator",
      rich_text = TRUE,
      hide_code = TRUE)
```

@@

These steps are summarized in Table \@ref(tab:table-summary). We will first demonstrate the ideas behind the steps using open data [@brysbaert2014; @balota2007]. This example will reveal a few areas of needed exploration for the steps. Next, we portray simulations for the proposed procedure and find solutions to streamline and improve the sample size estimation procedure. Table \@ref(tab:table-summary) shows the results of the simulations and solutions on the right hand side. Finally, we include additional resources for researchers to use to implement the estimation procedure.

@@

# grab only the overlap for this fake study
elpDF <- elpDF %>% 
  filter(Stimulus %in% conDF$Word) %>% 
  filter(Type == 1)

@@

```{r con-sample, include = F, echo = F}
# sequence of sample sizes to try
nsim <- 5
samplesize_values <- seq(20, 300, 5)
# create a blank table for us to save the values in 
sim_table <- matrix(NA, 
          nrow = length(samplesize_values)*nsim, 
          ncol = length(unique(con_use$Word)))
# make it a data frame
sim_table <- as.data.frame(sim_table)
# add a place for sample size values 
sim_table$sample_size <- NA

iterate <- 1
for (p in 1:nsim){
  # loop over sample sizes
  for (i in 1:length(samplesize_values)){

@@

   colnames(sim_table)[1:length(unique(con_use$Word))] <- temp$Word
   sim_table[iterate, 1:length(unique(con_use$Word))] <- temp$se
   sim_table[iterate, "sample_size"] <- samplesize_values[i]
   sim_table[iterate, "nsim"] <- p

   iterate <- iterate + 1
  }
}

# figure out potential samples
final_sample_c <- 
 sim_table %>% 
 pivot_longer(cols = -c(sample_size, nsim)) %>% 
 rename(item = name, se = value) %>% 
 group_by(sample_size, nsim) %>% 
 summarize(Percent_Below10 = sum(se <= cutoff_c["10%"])/length(unique(con_use$Word)),
          Percent_Below20 = sum(se <= cutoff_c["20%"])/length(unique(con_use$Word)),
          Percent_Below30 = sum(se <= cutoff_c["30%"])/length(unique(con_use$Word)),
          Percent_Below40 = sum(se <= cutoff_c["40%"])/length(unique(con_use$Word)),
          Percent_Below50 = sum(se <= cutoff_c["50%"])/length(unique(con_use$Word)),
          Percent_Below60 = sum(se <= cutoff_c["60%"])/length(unique(con_use$Word)),
          Percent_Below70 = sum(se <= cutoff_c["70%"])/length(unique(con_use$Word)),
          Percent_Below80 = sum(se <= cutoff_c["80%"])/length(unique(con_use$Word)),
          Percent_Below90 = sum(se <= cutoff_c["90%"])/length(unique(con_use$Word))) %>% 
  ungroup() %>% 
  group_by(sample_size) %>% 
  summarize(Percent_Below10 = mean(Percent_Below10),
            Percent_Below20 = mean(Percent_Below20),
            Percent_Below30 = mean(Percent_Below30),
            Percent_Below40 = mean(Percent_Below40),
            Percent_Below50 = mean(Percent_Below50),
            Percent_Below60 = mean(Percent_Below60),
            Percent_Below70 = mean(Percent_Below70),
            Percent_Below80 = mean(Percent_Below80),
            Percent_Below90 = mean(Percent_Below90)) %>% 
  pivot_longer(-sample_size) %>% 
  filter(value >= .80) %>% 
  group_by(name) %>% 
  arrange(sample_size, value)

@@

```{r elp-sample, include = F, echo = F}
# sequence of sample sizes to try
nsim <- 5
samplesize_values <- seq(20, 300, 5)
# create a blank table for us to save the values in 
sim_table <- matrix(NA, 
          nrow = length(samplesize_values), 
          ncol = length(unique(elp_use$Stimulus)))
# make it a data frame
sim_table <- as.data.frame(sim_table)
# add a place for sample size values 
sim_table$sample_size <- NA

iterate <- 1

for (p in 1:nsim){

  # loop over sample sizes
  for (i in 1:length(samplesize_values)){

@@

   colnames(sim_table)[1:length(unique(elp_use$Stimulus))] <- temp$Stimulus
   sim_table[iterate, 1:length(unique(elp_use$Stimulus))] <- temp$se
   sim_table[iterate, "sample_size"] <- samplesize_values[i]
   sim_table[iterate, "nsim"] <- p

   iterate <- iterate + 1
   }

}
# figure out potential samples
final_sample_e <- 
 sim_table %>% 
 pivot_longer(cols = -c(sample_size)) %>% 
 rename(item = name, se = value) %>% 
 group_by(sample_size, nsim) %>% 
 summarize(Percent_Below10 = sum(se <= cutoff_e["10%"])/length(unique(elp_use$Stimulus)),
          Percent_Below20 = sum(se <= cutoff_e["20%"])/length(unique(elp_use$Stimulus)),
          Percent_Below30 = sum(se <= cutoff_e["30%"])/length(unique(elp_use$Stimulus)),
          Percent_Below40 = sum(se <= cutoff_e["40%"])/length(unique(elp_use$Stimulus)),
          Percent_Below50 = sum(se <= cutoff_e["50%"])/length(unique(elp_use$Stimulus)),
          Percent_Below60 = sum(se <= cutoff_e["60%"])/length(unique(elp_use$Stimulus)),
          Percent_Below70 = sum(se <= cutoff_e["70%"])/length(unique(elp_use$Stimulus)),
          Percent_Below80 = sum(se <= cutoff_e["80%"])/length(unique(elp_use$Stimulus)),
          Percent_Below90 = sum(se <= cutoff_e["90%"])/length(unique(elp_use$Stimulus))) %>% 
  ungroup() %>% 
  group_by(sample_size) %>% 
  summarize(Percent_Below10 = mean(Percent_Below10),
            Percent_Below20 = mean(Percent_Below20),
            Percent_Below30 = mean(Percent_Below30),
            Percent_Below40 = mean(Percent_Below40),
            Percent_Below50 = mean(Percent_Below50),
            Percent_Below60 = mean(Percent_Below60),
            Percent_Below70 = mean(Percent_Below70),
            Percent_Below80 = mean(Percent_Below80),
            Percent_Below90 = mean(Percent_Below90)) %>% 
  pivot_longer(-sample_size) %>% 
  filter(value >= .80) %>% 
  group_by(name) %>% 
  arrange(sample_size, value)

@@

results_table_example <- as.data.frame(results_table)
colnames(results_table_example) <- c("Deciles", "C SE", "C 80", "C 85", "C 90", "C 95",
             "L SE", "L 80", "L 85", "L 90", "L 95")
results_table_example$`C Decile SE` <- apa_num(as.numeric(results_table_example$`C Decile SE`), digits = 2)
results_table_example$`L Decile SE` <- apa_num(as.numeric(results_table_example$`L Decile SE`), digits = 2)

apa_table(results_table_example, 
          caption = "Sample Size Estimates by Decile for Example Study",
          note = "C = Concreteness rating, L = Lexical Decision Response Latencies. Estimates are based on meeting at least the minimum percent of items (e.g., 80%) but may be estimated over that amount (e.g., 82.5%). SE columns represent the standard error value cutoff for each decile, while 80/85/90/95 percent columns represent the sample size needed to have that percent of items below the SE cutoff. For example, 100 participants are required to ensure at least 95% of concreteness items SE are below the 10 percent decile SE cutoff, and 175 participants are necessary for the lexical decision SE to be below its 10 percent decile cutoff.")
```

@@

```{r scale-size-figure, echo = F, fig.cap="Simulated pilot sample size and final projected sample size to achieve 80%, 85%, 90%, and 95% of items below threshold. These values are averaged over all other variables including decile. Black dots represent original sample size for reference. "}

percent_var <- sqrt(((100-0)^2)/4)
millisecond_var <- sqrt(((3000-0)^2)/4)

@@

```{r scale-skew-figure, echo = F, fig.cap="Simulated pilot sample size and final projected sample size to achieve 80%, 85%, 90%, and 95% of items below threshold. In comparison to Figure 1, this figure shows projected sample size for ceiling versus normal distributions on each scale. All other variables are averaged together, and black dots represent original sample size for reference." }

ggplot(summary_long %>% 
         group_by(original_n, power, data_type) %>% 
         summarize(sample_size_avg = mean(sample_size)),
       aes(original_n, sample_size_avg, color = data_type)) + 
  geom_point() + 
  geom_point(aes(original_n, original_n), color = "black") + 
  geom_line() + 
  geom_line(aes(original_n, original_n), color = "black") + 
  theme_linedraw(base_size = 15) + 
  xlab("\nPilot Sample Size") + 
  ylab("Projected Sample Size\n") + 
  scale_color_manual(name = "Scale Skew",
                     values = c(plot_colors)) +
  facet_wrap(~power,
             labeller = labeller(power = power.labs))+
  theme_bw()
```

@@

```{r scale-hetero-figure, echo = F, fig.cap="Simulated pilot sample size and final projected sample size to achieve 80%, 85%, 90%, and 95% of items below threshold. In comparison to Figure 1 and 2, this figure shows projected sample size or differing amounts of heterogeneity on each scale. All other variables are averaged together, and black dots represent original sample size for reference."}

ggplot(summary_long %>% 
         group_by(original_n, power, variability) %>% 
         summarize(sample_size_avg = mean(sample_size)),
       aes(original_n, sample_size_avg, color = variability)) + 
  geom_point() + 
  geom_point(aes(original_n, original_n), color = "black") + 
  geom_line() + 
  theme_linedraw() + 
  xlab("Pilot Sample Size") + 
  ylab("Projected Sample Size") + 
  scale_color_discrete(name = "Item Heterogeneity") +
  facet_wrap(~power)
```

@@

```{r corrected-figure, echo = F, fig.cap="Corrected projected sample sizes for variability and power levels to achieve 80%, 85%, 90%, and 95% of items below threshold. All other variables are averaged together, and black dots represent original sample size for reference."}

decay <- 1-sqrt((summary_long$original_n-20)/summary_long$original_n)^log2(summary_long$original_n)
summary_long$new_sample <- summary_long$sample_size*decay
ggplot(summary_long %>% 
         group_by(original_n, power, variability) %>% 
         summarize(sample_size_avg = mean(sample_size),
                   new_sample_size_avg = mean(new_sample)),
       aes(original_n, new_sample_size_avg, color = variability)) + 
  geom_line(aes(original_n, original_n), color = "black") +
  geom_point(aes(original_n, original_n), color = "black") + 
  geom_point() + 
  geom_line() + 
  xlab("\nPilot Sample Size") + 
  ylab("Projected Sample Size\n") + 
  scale_color_manual(name = "Item Heterogeneity",
                     values = c(plot_colors)) +
  facet_wrap(~power,
             labeller = labeller(power = power.labs))+
  theme_bw()
```

@@

The updated proposal steps are in Table \@ref(tab:table-summary) on the right hand side. The main change occurs in Step 2 with a designated cutoff decile, and Step 5 with a correction score. Using the data from the 40% decile in Table \@ref(tab:table-example), we can determine that the stopping rule for concreteness ratings would be `r apa_num(cutoff_c[4])`, and the stopping rule for lexical decision times would be `r apa_num(cutoff_e[4])`. For Step 5, we apply our correction formula separately for each one, as they have different variability scores, and these scores are shown in Table \@ref(tab:table-scores-updated). Each row was multiplied by row one's formula, and then these scores are summed for the final sample size. Sample sizes cannot be proportional, so we recommend rounding up to the nearest whole number.

@@

```{r boot-sample, include = T, echo = T}
samples <- bootstrap_samples(start = 20, # starting sample size
  stop = 100, # stopping sample size
  increase = 5, # increase bootstrapped samples by this amount
  population = pops, # population or pilot data
  replace = TRUE, # bootstrap with replacement? 
  nsim = 5, # number of simulations to run
  grouping_items = "item") # item column label
